Gap formation and stability in non-isothermal protoplanetary discs

### 2014-11-21 paper/summary.tex
penultimate edit by MKL

\section{Summary and discussion}\label{summary}
In this paper, we have carried out numerical simulations of non-isothermal disc-planet interaction. 
Our simulations were customised to examine the effect of a finite cooling time on the stability of gaps 
opened by giant planets to the so-called vortex or Rossby wave instability. To do so, we 
included an energy equation with a cooling term that restores the initial disc temperature on a characteristic timescale $t_c$. 
We studied the evolution of the gap stability as a function of $t_c$. This is a natural extension to previous
studies of on gap stability, which employ locally or strictly isothermal equations of state.

@@

In the second set of calculations, we included the planet potential throughout the simulations 
and examined the long-term evolution of the gap-edge vortex that develops from the RWI. The vortices to reach a
quasi-steady state lasting $O(10^3)$ orbits, during which its amplitude grows until it begins to induce shocks,  
afterwhich the vortex amplitudes decay. 
We observe vortex lifetimes to increase with increasing cooling time until
a cooling timescale on the order of the libration time at the gap edge {\bf (this statement assumes longest lifetime is $\tilde{\beta}=1$, so may change
depending on new definitions of lifetimes)}, beyond which vortex lifetimes shorten again.

Our results indicate there exists an optimum cooling timescale to maximise the lifetime of gap-edge vortices. We suggest the non-monotonic dependence
can be attributed to the time required for the vortex to grow to sufficient amplitude to induce shocks in the surrounding fluid, thereby losing energy and also smooth out the gap edge.  
For short cooling timescales, the planet is able to open a deeper gap which favours the RWI, leading to stronger vortices. However, for long 
cooling timescales, the vortex grow faster during the quasi-steady state {\bf check if statement true}. 
We suggest the latter is due to the RWI being favoured with increasing disc temperature, according to previous stability
calculations. These competing factors imply for both very short and long cooling timescales, the vortex reaches its maximum amplitude 
, shocks, and begins to decay, sooner than intermediate cooling timescales. 
%decay timescale depends on cooling time, but there seems to be a monotonic trend: longer cooling times, decay timescale is shorter, so this 
%doesn't contribute to non-monotonic part

We remark that a non-monotonic dependence of the vortex lifetime was also reported
by \cite{fu14}, who performed isothermal disc-planet simluations with different values of the 
disc aspect ratio. In their simulations the optimum aspect ratio is $h=0.06$. In our simulations,
the aspect ratio is a dynamical variable, but by analysing the region where the vortex is located ($r-r_p\in[2,10]r_h$), 
% {\bf define this region}
we find for $\tilde\beta=5.0$ case, with the longest vortex lifetime, has
$h\approx0.06$ on average throughout its lifetime. 
{\bf if new definition of vortex lifetime gives other $\tilde{\beta}$ values as the maximum lifetime, then need to update this value of $h$}. 
Our result is therefore qualitatively consistent with \citep{fu14}. 
%This value of $h$ was also
%found to maximise vortex lifetimes in the simulations of
%\cite{fu14}. Note that $h$ is a fixed parameter in \cite{fu14} while in
%our simulations $h$ is a dynamical variable.

@@

\emph{Convergence.} Although lower-resolution simulations performed in the early stages of this project gave similar results 
(most importantly, the non-monotonic dependence of vortex lifetimes on the cooling timescale), we did find the lower-resolution simulations 
typically yield longer vortex lifetimes than that reported in this paper. This could be due to weaker RWI with low resolution. 
It will be neccessary to perform even higher resolution  simulations in order to obtain quantitatively converged vortex lifetimes.

@@

\emph{Three-dimensional (3D) effects.} A vortex in a 3D disc may be subject to secondary instabilities 
that destroy them \citep{lesur09,railton14}. This may an important factor in determining gap-edge vortex lifetimes
in realistic discs. For example, if these secondary instabilities sets in before the vortex grows to sufficient amplitude to shock, 
then the dependence of the vortex lifetime on the cooling timescale will be that through the 3D instability (as opposed to the effect on the RWI itself).  
This problem needs to be clarified with full 3D disc-planet simulations.

>>>>

\section{Summary and discussion}\label{summary}
In this paper, we have carried out numerical simulations of
non-isothermal disc-planet interaction.  
Our simulations were customised to examine the effect of a finite 
cooling time on the stability of gaps  
opened by giant planets to the so-called vortex or Rossby wave
instability. To do so, we  
included an energy equation with a cooling term that restores the 
disc temperature to its initial profile on a characteristic timescale
$t_c$. We studied the evolution of the gap stability as a function of 
$t_c$. This is a natural extension to previous studies of on gap
stability, which employ locally or strictly isothermal equations of
state.

@@

In the second set of calculations, we included the planet potential
throughout the simulations  and examined the long-term evolution of
the gap-edge vortex that develops from the RWI. The vortex reaches 
a quasi-steady state lasting $O(10^3)$ orbits. Unlike the `planet-off'
simulations, in which vortices decay after linear growth and merging,
we find that with the planet potential kept on, the vortex amplitude
grows during this quasi-steady state, until it begins to induce
shocks, afterwhich the vortex amplitude begins to decay.

We observe the duration of the quasi-steady state increases with
increasing cooling time until some critical value, beyond which this
quasi-steady state shortens again. The timescale for the vortex to
decay, after reaching maximum amplitude, decreases monotonically with
increasing cooling time.

%technically we should probably say there is an optimum cooling time
%to maximise the quasi-steady state

Our results indicate there exists an optimum cooling timescale to
maximise the lifetime of gap-edge vortices. We suggest the
non-monotonic dependence of the quasi-steady state can be attributed
to the time required for the vortex to grow to sufficient amplitude to
induce shocks in the surrounding fluid, thereby losing energy and also
smooth out the gap edge.

For short cooling timescales, the planet is able to open a
deeper gap which favours the RWI, leading to stronger
vortices. However, for long  cooling timescales, the vortex grow
faster during the quasi-steady state {\bf check if statement true}. We
suggest the latter is due to the RWI being favoured with increasing
disc temperature, according to previous stability calculations. These
competing factors imply for both very short and long cooling
timescales, the vortex reaches its maximum amplitude, shock, and
begins to decay, sooner than intermediate cooling timescales.

%decay timescale depends on cooling time, but there seems to be a
%monotonic trend: longer cooling times, decay timescale is shorter, so
%this  
%doesn't contribute to non-monotonic trend

We remark that a non-monotonic dependence of the vortex lifetime was
also reported by \cite{fu14}, who performed isothermal disc-planet
simluations with different values of the   
disc aspect ratio. In their simulations the optimum aspect ratio is
$h=0.06$. In our simulations, the disc aspect ratio is a dynamical
variable, but by analysing the region where the vortex is located
($r-r_p\in[2,10]r_h$), % {\bf define this region}
we find for $\tilde\beta=5.0$ case, with the longest vortex lifetime, has
$h\approx0.06$ on average throughout its lifetime. 
{\bf if new definition of vortex lifetime gives other $\tilde{\beta}$ values as the maximum lifetime, then need to update this value of $h$}. 
Our result is therefore consistent with \citep{fu14}.

% we don't know how fu defined vortex lifetime

@@

\emph{Convergence.} Although lower-resolution simulations performed in
the early stages of this project gave similar results  
(most importantly, the non-monotonic dependence of vortex lifetimes on
the cooling timescale), we did find the lower-resolution simulations  
typically yield longer vortex lifetimes than that reported in this
paper. This could be due to weaker RWI with low resolution.  
It will be neccessary to perform even higher resolution  simulations
in order to obtain quantitatively co nverged vortex lifetimes.

@@

\emph{Three-dimensional (3D) effects.} A vortex in a 3D disc may be
subject to secondary instabilities that destroy them
\citep{lesur09,railton14}. This may be an important factor in
determining gap-edge vortex lifetimes in realistic discs. For example,
if these secondary instabilities sets in before the vortex grows to
sufficient amplitude to shock, then the dependence of the vortex
lifetime on the cooling timescale will be its effect through the 3D
instability (as opposed to the effect on the RWI itself, which is a 2D
instability). This problem needs to be clarified with full 3D
disc-planet simulations.
